### 2019-04-01 Thesis/games.tex
D.

\begin{figure}
  \begin{mathpar}
    \Gamma,s\ts (\lambda x.M)N \oopto \Gamma,s\ts M[N/x]
    \and
    \Gamma,s\ts \Y M \oopto \Gamma,s\ts M(\Y M)
    \and
    \Gamma,s\ts \suc n \oopto \Gamma,s\ts n+1
    \\
    \Gamma,s \ts \pred (n+1) \oopto \Gamma,s\ts n
    \and
    \Gamma,s\ts \pred 0 \oopto \Gamma,s\ts 0
    \\
    \Gamma,s\ts \skipp;M \oopto \Gamma,s\ts M
    \\
    \Gamma,s\ts \If \true \Then N \Else P \oopto \Gamma,s\ts N
    \and
    \Gamma,s\ts \If \false \Then N \Else P \oopto \Gamma,s\ts P
    \and
    \Gamma,s\ts \IfO 0 \Then N \Else P \oopto \Gamma,s\ts N
    \and
    \Gamma,s\ts \IfO (n+1) \Then N \Else P \oopto \Gamma,s\ts P
    \and
    \Gamma,s\ts x \gets n \oopto \Gamma,(s\vert x\mapsto n)\ts \skipp
    \and
    \Gamma,s\ts \oc x \oopto \Gamma,s\ts s(x)
    \\
    \Gamma,s\ts \neww \lambda x.M \oopto \Gamma,x,(s\vert x\mapsto 0)\ts M
    \\
    \Gamma,s \ts (\mkvar W\,R) \gets n \oopto \Gamma,s\ts Wn
    \and
    \Gamma,s \ts \oc(\mkvar W\,R) \oopto \Gamma,s\ts R
  \end{mathpar}
  \caption{Felleisen-style small-step operational semantics for Idealized Algol.}
  \label{FigIaSsOpSem}
\end{figure}

@@

\begin{lemma}
  Suppose that $\Gamma,s\ts M \oopto \Gamma,\Delta,s'\ts N$, where $M,N$ have type $T$.
  Then the following diagram commutes.
  \[
    \begin{tikzcd}[column sep=28pt]
      I \arrow[r, "\oc \deno s"] \arrow[d, "\oc \deno s'"']
        & \oc S_\Gamma \arrow[r, "\cell^\Gamma"]
          & \oc \V_\Gamma \arrow[r, "\mu_{\V_\Gamma}"]
            &[22pt] \oc \V_\Gamma \tensor \oc \V_\Gamma \arrow[d, "\deno{\Gamma\ts M}\tensor \oc V_\Gamma"] \\
      {\oc S_{\Gamma,\Delta}} \arrow[d, "{\cell^{\Gamma,\Delta}}"']
        &
          &
            & \deno T \tensor \oc V_\Gamma \arrow[d, "{\wk_{\deno T,\oc \V_{\Gamma,\Delta}}}"] \\
      {\oc \V_{\Gamma,\Delta}} \arrow[d, "{\mu_{\V_{\Gamma,\Delta}}}"']
        &
          &
            &  \deno{T}\sequoid \oc\V_\Gamma \arrow[d, "{\deno{T} \sequoid (\Gamma+0_\Delta)}"] \\
      {\oc \V_{\Gamma,\Delta} \tensor \oc \V_{\Gamma,\Delta}} \arrow[rr, "{\deno{\Gamma,\Delta\ts N} \tensor \oc \V_{\Gamma,\Delta}}"]
        &
          & {\deno T \tensor \oc \V_{\Gamma,\Delta}} \arrow[r, "{\wk_{\deno T,\oc \V_{\Gamma,\Delta}}}"]
            & {{\deno T \sequoid \oc \V_{\Gamma,\Delta}}\,,}
    \end{tikzcd}
    \]
  where $\Gamma+0_\Delta\from \oc \V_\Gamma \to \oc \V_{\Gamma,\Delta}$ is the composite
  \[
    \oc \V_\Gamma \xrightarrow{\runit_{\oc\V_\Gamma}} \oc \V_\Gamma \tensor I \xrightarrow{\oc\V_\Gamma\tensor \oc0} \oc\V_\Gamma \tensor \oc S_\Delta \xrightarrow{\oc\V_\Gamma\tensor \cell^\Delta} \oc \V_\Gamma \tensor \oc \V_\Delta \to \oc \V_{\Gamma,\Delta}\,;
    \]
  i.e., the morphism given by initializing new storage cells to $0$ for all the new variables in $\Delta$.
  \label{LemSoundnessOopto}
\end{lemma}
\begin{proof}
  We prove this on a case-by-case basis.
  \begin{itemize}
    \item For most of the rules, $\Delta=\_$ and $s=s'$; i.e., the rule is of the form
      \[
        \Gamma,s\ts M \oopto \Gamma,s \ts N
        \]
      for some $M,N$.
      In such a case, it suffices to show that $\deno{\Gamma\ts M}=\deno{\Gamma\ts N}$.
      Indeed, we have
      \begin{itemize}
        \item $\deno{\Gamma\ts(\lambda x.M)N} = \deno{\Gamma\ts M[N/x]}$ (by a usual substitution-lemma argument);
        \item $\deno{\Gamma\ts\Y M} = \deno{\Gamma\ts (\lambda F.\lambda f.f(F f)) \Y M} = \deno{\Gamma\ts M(\Y M)}$;
        \item $\deno{\Gamma\ts \suc n} = \deno{\Gamma \ts n+1}$;
        \item $\deno{\Gamma\ts \pred (n+1)} = \deno{\Gamma\ts n}$;
        \item $\deno{\Gamma\ts \pred 0} = \deno{\Gamma\ts 0}$;
        \item $\deno{\Gamma\ts\skipp;M} = \deno{\Gamma\ts M}$;
        \item $\deno{\Gamma\ts\If \true \Then N \Else P} = \deno{\Gamma\ts (\lambda x.\lambda y.x)NP} = \deno{\Gamma\ts N}$;
        \item $\deno{\Gamma\ts\If \false \Then N \Else P} = \deno{\Gamma\ts (\lambda x.\lambda y.y)NP} = \deno{\Gamma\ts P}$;
        \item $\deno{\Gamma\ts\IfO 0 \Then N \Else P} = \deno{\Gamma\ts (\lambda x.\lambda y.x)NP} = \deno{\Gamma\ts N}$;
        \item $\deno{\Gamma\ts\IfO (n+1) \Then N \Else P} = \deno{\Gamma\ts (\lambda x.\lambda y.y)NP} = \deno{\Gamma\ts P}$;
        \item \parbox[t][][t]{0.8\textwidth}{$\deno{\Gamma\ts(\mkvar W\,R)\gets n} = \langle \deno{\Gamma\ts Wn}_{n\in\bN},\deno{\Gamma\ts R}\rangle;\pr_n = \deno{\Gamma\ts Wn}$; and}
        \item $\deno{\Gamma\ts\oc(\mkvar W\,R)} = \langle \deno{\Gamma\ts Wn}_{n\in\bN},\deno{\Gamma\ts R}\rangle;\pr_{\bN} = \deno{\Gamma\ts R}$.
      \end{itemize}
    \item Now consider the rule
      \[
        \Gamma,s\ts\neww \lambda x.M\oopto \Gamma,x,(s\vert x\mapsto 0)\ts M\,,
        \]
      where $\Gamma\ts M\from T$.
      The first observation to make is that the denotation $\deno{\Gamma\ts \neww \lambda x.M}$ may be written as
      \[
        \oc\V_\Gamma \xrightarrow{\Gamma + 0_x} \oc\V_{\Gamma,x} \xrightarrow{\Lambda\inv(\deno{\Gamma \ts \lambda x.M})} \deno{T}\,,
        \]
      where $\Gamma+0_x$ is as defined in the statement of this Lemma.
      Of course, $\Lambda\inv(\deno{\Gamma\ts \lambda x.M})=\deno{\Gamma,x\ts M}$.

@@

@@

        \Gamma,s\ts x\gets n \oopto \Gamma,(s\vert x\mapsto n) \ts \skipp\,.

@@

>>>>

\begin{figure}
  \begin{mathpar}
    \Gamma,s\ts (\lambda x.M)N \oopto \Gamma,s\ts M[N/x]
    \and
    \Gamma,s\ts \Y M \oopto \Gamma,s\ts M(\Y M)
    \and
    \Gamma,s\ts \suc n \oopto \Gamma,s\ts n+1
    \\
    \Gamma,s \ts \pred (n+1) \oopto \Gamma,s\ts n
    \and
    \Gamma,s\ts \pred 0 \oopto \Gamma,s\ts 0
    \\
    \Gamma,s\ts \skipp;M \oopto \Gamma,s\ts M
    \\
    \Gamma,s\ts \If \true \Then N \Else P \oopto \Gamma,s\ts N
    \and
    \Gamma,s\ts \If \false \Then N \Else P \oopto \Gamma,s\ts P
    \and
    \Gamma,s\ts \IfO 0 \Then N \Else P \oopto \Gamma,s\ts N
    \and
    \Gamma,s\ts \IfO (n+1) \Then N \Else P \oopto \Gamma,s\ts P
    \and
    x,\Gamma,s\ts x \gets n \oopto x,\Gamma,(s\vert x\mapsto n)\ts \skipp
    \and
    x,\Gamma,s\ts \oc x \oopto x,\Gamma,s\ts s(x)
    \\
    \Gamma,s\ts \neww \lambda x.M \oopto \Gamma,x,(s\vert x\mapsto 0)\ts M
    \\
    \Gamma,s \ts (\mkvar W\,R) \gets n \oopto \Gamma,s\ts Wn
    \and
    \Gamma,s \ts \oc(\mkvar W\,R) \oopto \Gamma,s\ts R
  \end{mathpar}
  \caption{Felleisen-style small-step operational semantics for Idealized Algol.}
  \label{FigIaSsOpSem}
\end{figure}

@@

\begin{lemma}
  The relation $\oopto$ betwen triples $\Gamma,s\ts M$ preserves the composite
    \oc\deno{s};\cell^\Gamma;\mu_{\V_\Gamma};(\deno{\Gamma\ts M}\tensor\oc\V_\Gamma);\wk_{\deno T,\oc\V_{\Gamma}}\,.
  In other words, if $\Gamma,s\ts M \oopto \Gamma,\Delta,s'\ts N$, where $M,N$ have type $T$, then the following diagram commutes.
    \begin{tikzcd}[column sep=28pt]
      I \arrow[r, "\oc \deno s"] \arrow[d, "\oc \deno s'"']
        & \oc S_\Gamma \arrow[r, "\cell^\Gamma"]
          & \oc \V_\Gamma \arrow[r, "\mu_{\V_\Gamma}"]
            &[22pt] \oc \V_\Gamma \tensor \oc \V_\Gamma \arrow[d, "\deno{\Gamma\ts M}\tensor \oc V_\Gamma"] \\
      {\oc S_{\Gamma,\Delta}} \arrow[d, "{\cell^{\Gamma,\Delta}}"']
            & \deno T \tensor \oc V_\Gamma \arrow[d, "{\wk_{\deno T,\oc \V_{\Gamma}}}"] \\
      {\oc \V_{\Gamma,\Delta}} \arrow[d, "{\mu_{\V_{\Gamma,\Delta}}}"']
            &  \deno{T}\sequoid \oc\V_\Gamma \arrow[d, "{\deno{T} \sequoid (\Gamma+0_\Delta)}"] \\
      {\oc \V_{\Gamma,\Delta} \tensor \oc \V_{\Gamma,\Delta}} \arrow[rr, "{\deno{\Gamma,\Delta\ts N} \tensor \oc \V_{\Gamma,\Delta}}"]
          & {\deno T \tensor \oc \V_{\Gamma,\Delta}} \arrow[r, "{\wk_{\deno T,\oc \V_{\Gamma,\Delta}}}"]
            & {{\deno T \sequoid \oc \V_{\Gamma,\Delta}}\,,}
  where $\Gamma+0_\Delta\from \oc \V_\Gamma \to \oc \V_{\Gamma,\Delta}$ is the composite
    \oc \V_\Gamma \xrightarrow{\runit_{\oc\V_\Gamma}} \oc \V_\Gamma \tensor I \xrightarrow{\oc\V_\Gamma\tensor \oc0} \oc\V_\Gamma \tensor \oc S_\Delta \xrightarrow{\oc\V_\Gamma\tensor \cell^\Delta} \oc \V_\Gamma \tensor \oc \V_\Delta \to \oc \V_{\Gamma,\Delta}\,;
  i.e., the morphism given by initializing new storage cells to $0$ for all the new variables in $\Delta$.
  \label{LemSoundnessOopto}
\end{lemma}
\begin{proof}
  We prove this on a case-by-case basis.
  \begin{itemize}
    \item For most of the rules, $\Delta=\_$ and $s=s'$; i.e., the rule is of the form
        \Gamma,s\ts M \oopto \Gamma,s \ts N
      for some $M,N$.
      In such a case, it suffices to show that $\deno{\Gamma\ts M}=\deno{\Gamma\ts N}$.
      Indeed, we have
      \begin{itemize}
        \item $\deno{\Gamma\ts(\lambda x.M)N} = \deno{\Gamma\ts M[N/x]}$ (by a usual substitution-lemma argument);
        \item $\deno{\Gamma\ts\Y M} = \deno{\Gamma\ts (\lambda F.\lambda f.f(F f)) \Y M} = \deno{\Gamma\ts M(\Y M)}$;
        \item $\deno{\Gamma\ts \suc n} = \deno{\Gamma \ts n+1}$;
        \item $\deno{\Gamma\ts \pred (n+1)} = \deno{\Gamma\ts n}$;
        \item $\deno{\Gamma\ts \pred 0} = \deno{\Gamma\ts 0}$;
        \item $\deno{\Gamma\ts\skipp;M} = \deno{\Gamma\ts M}$;
        \item $\deno{\Gamma\ts\If \true \Then N \Else P} = \deno{\Gamma\ts (\lambda x.\lambda y.x)NP} = \deno{\Gamma\ts N}$;
        \item $\deno{\Gamma\ts\If \false \Then N \Else P} = \deno{\Gamma\ts (\lambda x.\lambda y.y)NP} = \deno{\Gamma\ts P}$;
        \item $\deno{\Gamma\ts\IfO 0 \Then N \Else P} = \deno{\Gamma\ts (\lambda x.\lambda y.x)NP} = \deno{\Gamma\ts N}$;
        \item $\deno{\Gamma\ts\IfO (n+1) \Then N \Else P} = \deno{\Gamma\ts (\lambda x.\lambda y.y)NP} = \deno{\Gamma\ts P}$;
        \item \parbox[t][][t]{0.8\textwidth}{$\deno{\Gamma\ts(\mkvar W\,R)\gets n} = \langle \deno{\Gamma\ts Wn}_{n\in\bN},\deno{\Gamma\ts R}\rangle;\pr_n = \deno{\Gamma\ts Wn}$; and}
        \item $\deno{\Gamma\ts\oc(\mkvar W\,R)} = \langle \deno{\Gamma\ts Wn}_{n\in\bN},\deno{\Gamma\ts R}\rangle;\pr_{\bN} = \deno{\Gamma\ts R}$.
        \Gamma,s\ts\neww \lambda x.M\oopto \Gamma,x,(s\vert x\mapsto 0)\ts M\,,
      where $\Gamma\ts M\from T$.
      The first observation to make is that the denotation $\deno{\Gamma\ts \neww \lambda x.M}$ may be written as
        \oc\V_\Gamma \xrightarrow{\Gamma + 0_x} \oc\V_{\Gamma,x} \xrightarrow{\Lambda\inv(\deno{\Gamma \ts \lambda x.M})} \deno{T}\,,
      where $\Gamma+0_x$ is as defined in the statement of this Lemma.
      Of course, $\Lambda\inv(\deno{\Gamma\ts \lambda x.M})=\deno{\Gamma,x\ts M}$.

@@

    \item Now consider the rules
        x,\Gamma,s\ts x\gets n \oopto x,\Gamma,(s\vert x\mapsto n) \ts \skipp
        x,\Gamma,s\ts \oc x\oopto x,\Gamma,s\ts s(x)\,.
      In each of these cases, the outer context $\Gamma$ is unchanged by the rule; moreover, each term ignores the variables in $\Gamma$.
      So the denotations terms-in-context $x,\Gamma\ts M\from T$ on either side of each rule are of the form

@@

        \oc\V_{x,\Gamma} \xrightarrow{\oc\pr_1} \oc \Varr \xrightarrow{\deno{x\ts M}} \deno{T}\,.

@@

      Now the diagram in Figure \ref{FigSoundnessOoptoContext} tells us that each
        \oc\deno{s};\cell^{x,\Gamma};\mu_{\V_{x,\Gamma}};(\deno{x,\Gamma\ts M}\tensor\oc\V_{x,\Gamma});\wk_{\deno T,\oc\V_{x,\Gamma}}\,\text{--}
      i.e., the composites we are trying to show are preserved by the $\oopto$ relation -- may be written as
        \oc\deno{s\vert_\Gamma};\cell^\Gamma;\lunit_{\oc\V_\Gamma};\\((\oc\deno{s\vert_x};\cell;\mu_{\Varr};(\deno{x\ts M}\tensor\oc\Varr);\wk_{\deno T,\oc\Varr})\tensor\oc\V_\Gamma);\\\wk_{\deno T\sequoid \oc\Varr,\oc\V_\Gamma};\passoc_{\deno T,\oc\Varr,\oc\V_\Gamma};\cong
      (where $\cong$ represents the natural isomorpism between $\deno T\sequoid(\oc\Varr\tensor\oc\V_\Gamma)$ and $\deno T \sequoid\oc\V_{x,\Gamma}$),
      and so is determined by the composite
        \oc\deno{s\vert_x};\cell;\mu_{\Varr};(\deno{x\ts M}\tensor\oc\Varr);\wk_{\deno T,\oc\Varr}\,;
      i.e., the composite that we would get if $\Gamma$ were empty.
      This tells us that we can assume that $\Gamma$ is the empty context when we want to deal with these two cases.
          \begin{tikzcd}[ampersand replacement=\&, column sep=5pt]
            I \arrow[r, "\oc\deno{s\vert_\Gamma}"] \arrow[d, "\oc\deno s"' {description, yshift=2pt}, thick, dashed]
              \& \oc S_\Gamma \arrow[r, "\cell^\Gamma"] \arrow[dr, "\lunit_{\oc S_\Gamma}" description, dotted]
                \&[18pt] \oc \V_\Gamma \arrow[r, "\lunit_{\oc\V_\Gamma}"] 
                  \&[18pt] I \tensor \oc\V_\Gamma \arrow[dd, "\oc\deno{s\vert_x}\tensor\oc\V_\Gamma" description, thick] \\
            {\oc S_{x,\Gamma}} \arrow[drr, Isom, dotted] \arrow[dd, "{\cell^{x,\Gamma}}"' description, thick, dashed]
                \& I \tensor \oc S_\Gamma \arrow[ur, "I \tensor\cell^\Gamma" description, dotted] \arrow[d, "\oc\deno{s\vert_x}\tensor\oc S_\Gamma" description, dotted]
                  \& \\
                \& \oc\bN\tensor\oc S_\Gamma \arrow[r, "\oc\bN\tensor\cell^\Gamma" description, dotted] \arrow[dr, "\cell\tensor\cell^\Gamma" description, dotted]
                  \& \oc\bN \tensor\oc\V_\Gamma \arrow[d, "\cell\tensor\oc\V_\Gamma" description, thick] \\
            {\oc\V_{x,\Gamma}} \arrow[rrr, Isom, dotted] \arrow[d, "{\mu_{\V_{x,\Gamma}}}"' description, thick, dashed]
                  \& \oc\Varr \tensor \oc\V_\Gamma \arrow[d, "\mu_{\Varr}\tensor\oc\V_\Gamma" description, thick] \arrow[dl, "\mu_{\Varr}\tensor\mu_{\V_\Gamma}" description, bend right=15, dotted] \\[10pt]
            {\oc\V_{x,\Gamma}\tensor\oc\V_{x,\Gamma}} \arrow[r, dotted] \arrow[d, "{\oc\pr_1\tensor\oc\V_{x,\Gamma}}" description, dotted] \arrow[dd, "{\deno{x,\Gamma\ts M}\tensor\oc\V_{x,\Gamma}}" description, bend right=68, thick, dashed, end anchor={[xshift=-3pt]}]
              \& (\oc\Varr \tensor \oc\V_\Gamma) \tensor (\oc\Varr \tensor\oc\V_\Gamma) \arrow[r, dotted] \arrow[d, "(\oc\Varr\tensor())\tensor(\oc\Varr\tensor\oc\V_{\Gamma})" description, dotted]
                \& (\oc\Varr \tensor \oc\Varr) \tensor (\oc\V_\Gamma \tensor\oc\V_\Gamma) \arrow[r, "(\oc\Varr\tensor\oc\Varr)\tensor(()\tensor\oc\V_\Gamma)" yshift=5pt, dotted]
                  \& (\oc\Varr \tensor \oc\Varr) \tensor\oc\V_\Gamma \arrow[d, "(\deno{x\ts M}\tensor \oc\Varr) \tensor \oc\V_\Gamma" description, thick] \\[30pt]
            {\oc\Varr \tensor \oc\V_{x,\Gamma}} \arrow[r, Isom, dotted] \arrow[d, "{\deno{x\ts M} \tensor \oc\V_{x,\Gamma}}"' {description, yshift=1pt}, dotted]
              \& \oc\Varr \tensor (\oc\Varr \tensor\oc\V_\Gamma) \arrow[urr, "{\assoc_{\oc\Varr,\oc\Varr,\oc\V_\Gamma}\inv}" {description, xshift=-5pt, yshift=5pt}, start anchor={[xshift=-25pt, dotted]}, bend left=5] \arrow[r, "\deno{x\ts M} \tensor (\oc\Varr\tensor\oc\V_\Gamma)" yshift=2pt, dotted]
                \& \deno{T} \tensor (\oc\Varr \tensor \oc\V_\Gamma) \arrow[r, "{\assoc_{\deno T,\oc\Varr,\oc\V_\Gamma}}" yshift=2pt, dotted] \arrow[ddl, "{\wk_{\deno T,(\oc\Varr\tensor\oc\V_\Gamma}}" description, dotted]
                  \& (\deno T \tensor \oc\Varr) \tensor \oc\V_\Gamma\arrow[d, "{\wk_{\deno T,\oc\Varr}\tensor\oc\V_\Gamma}" {description, yshift=1pt}, thick] \\
            {\deno T \tensor \oc\V_{x,\Gamma}} \arrow[d, "{\wk_{\deno T,\oc\V_{x,\Gamma}}}"' {description, yshift=2pt}, thick, dashed] \arrow[urr, Isom, dotted]
                  \& (\deno T \sequoid \oc\Varr) \tensor \oc\V_\Gamma \arrow[d, "{\wk_{(\deno T\sequoid\oc\Varr),\oc\V_\Gamma}}" {description, yshift=2pt}] \\[5pt]
            {\deno T \sequoid \oc\V_{x,\Gamma}} \arrow[r, Isom]
              \& \deno T \sequoid (\oc\Varr \tensor\oc\V_\Gamma)
                  \& (\deno T \sequoid \oc\Varr) \sequoid\oc\V_\Gamma \arrow[ll, "{\passoc_{\deno T,\oc\Varr,\oc\V_\Gamma}}"]
          \caption{Diagram proving that if we want to prove the conclusion of Lemma \ref{LemSoundnessOopto} for a small-step rule that does not change the context and only mentions one variable, then it suffices to assume that that variable is the only variable in the context.}
          \label{FigSoundnessOoptoContext}
